---
title: "Week 8: Exercises"
output: github_document
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(readr)
library(forcats)
library(ggplot2)
library(tidyr)
library(stringr)
```

## Exercise: 

```{r}
ski_acres <- data.frame( 
  ski.hill = c('Big Sky','Bridger Bowl', 'Jackson', 'Steamboat' ), 
                        skiable.acres = c(5800,2000, "2500+",2965))
```


### Q1. 
What type of variable is `skiable.acres` in the `ski_acres`
data frame?

### Q2.
Sort the ski resorts by skiable acres in descending fashion.


## Exercise

Now combine the following information into a single table sorted alphabetically by the name of the ski hill.
```{r, echo = F}
ski_acres
df_cost <- data.frame( ski.resort = c('Bridger Bowl', 'Big Sky', 'Steamboat', 'Jackson'), 
                       ticket.cost = c(60, 'depends',145, 130))
df_cost
disco <- c('Discovery', 2200, 20); disco
```

## Exercise

Combine the two data sets using join
```{r,echo = F}
ski_acres
df_cost
disco_df <- tibble(ski.hill = 'Discovery', skiable.acres = '2200', ticket.cost = '20'); disco_df
```

## Exercise

```{r}
billboard
```

Determine which song in this dataset spent the most time at #1

## Exercise

Using either the Seattle Housing dataset or the billboard data, explore the packages above.

```{r, echo = F}
seattle_in <- read_csv('http://math.montana.edu/ahoegh/teaching/stat408/datasets/SeattleHousing.csv')

billboard_thin <- billboard %>% 
  pivot_longer(
  cols= starts_with('wk'),
  names_to = 'week',
  values_to = 'rank',
  values_drop_na = T) %>%
  mutate(week_numb= as.numeric(str_replace(week, 'wk',''))) %>% mutate(date.of.rank = date.entered + (7 * (week_numb - 1)))

```

