---
title: "Week 8: Merging, Maps, and More"
output:
  revealjs::revealjs_presentation:
    theme: white
    center: true
    transition: none
    incremental: true
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(readr)
library(forcats)
library(ggplot2)
library(tidyr)
library(stringr)
```

# Factors

## Factors...
Factors are a specific way to store categorical data. Usign factors results in a more efficient data storage process, but can be cumbersome.

Factors can be necessary for making plots and fitting models in R.

## Character Values

```{r}
favorite_day <- c('Friday', 'Saturday', 'Sunday')
class(favorite_day)
sort(favorite_day)
```

## Creating Factors

```{r}
favorite_day <- c('Friday', 'Saturday', 'Sunday', 'Monday')
sort(favorite_day)
day_factor <- as.factor(favorite_day)
class(day_factor)
sort(day_factor)
```

## Creating Factors
Rather than coercing a class variable to be a factor, the factor can be created directly.
```{r}
day_factor2 <- factor(c('Friday', 'Saturday', 'Sunday', 'Monday'), 
levels = c('Sunday','Monday','Tuesday','Wednesday','Thursday','Friday','Saturday'))
sort(day_factor2)
```


# Sort / Order / Arrange

## Sort
We have seen the sort command, this returns a sorted result.
```{r}
ski_acres <- data.frame( ski.hill = c('Big Sky','Bridger Bowl', 'Jackson', 'Steamboat' ), 
                        skiable.acres = c(5800,2000, "2500+",2965))
kable(ski_acres)
```

## Exercise: 

### Q1. 
What type of variable is `skiable.acres` in the `ski_acres`
data frame?

### Q2.
Sort the ski resorts by skiable acres in descending fashion.


## Solution: 

### Q1. 
What type of variable is `skiable.acres` in the `ski_acres`
data frame?
```{r}
class(ski_acres$skiable.acres)
```

## Solution: 


### Q2.
Sort the ski resorts by skiable acres in descending fashion.
```{r}
ski_acres %>% arrange(desc(skiable.acres))
```

## arrange

Arrange is one of the verb words in the `dplyr()` package and can be very useful for arranging data frames.

## sort/order

Otherwise, the `sort` and `order` functions, in base R, can be used.

```{r}
order(ski_acres$skiable.acres)
ski_acres[order(ski_acres$skiable.acres, decreasing = T),]
```

## categorical variables and factors

We have already seen sorting on factor variables, what about plain categorical variables?

```{r}
sort(favorite_day)

string_vals <- c('1','10','2','3')

sort(string_vals)
```


# Merge / Join

## Merge
Another important skill is merging or combining data sets.

Consider the two data frames, how can we merge them and what should be the dimensions of the merged data frame.
```{r, echo = FALSE}
df1 <- data.frame(school = c('MSU','VT','Mines'),
            state= c('MT','VA','CO'), stringsAsFactors = F)
kable(df1)
df2 <- data.frame(school = c('Mines','MSU','VT'),
            enrollment = c(5794,15688,30598), stringsAsFactors = F)
kable(df2)
```

## pre-sort
One possibility is to use the arrange the data frames first.

```{r}
df1 <- df1[order(df1$school),]
df2 <- df2[order(df2$school),]
```

## pre-sort
One possibility is to use the arrange the data frames first.

```{r}
df1 
df2 
```


## rbind() and cbind()

Now, given that the data frames are both sorted the same way, we can bind the rows together.

```{r}
comb_df <- cbind(df1,df2)
comb_df
comb_df <- comb_df[,-3]
```

## rbind() and cbind()
Now assume we want to add another school to the data frame.

```{r, error=TRUE}
new.school <- c('Luther', 'IA',2337)
rbind(comb_df, new.school)
```
Note: if your strings are saved as factors, this chunk of code will give you an error.

## bind_rows() / bind_cols()

`dplyr` also contains functions for binding rows and columns.

## Exercise

Now combine the following information into a single table sorted alphabetically by the name of the ski hill.
```{r, echo = F}
ski_acres
df_cost <- data.frame( ski.resort = c('Bridger Bowl', 'Big Sky', 'Steamboat', 'Jackson'), 
                       ticket.cost = c(60, 'depends',145, 130))
df_cost
disco <- c('Discovery', 2200, 20); disco
```

## Solution part 1

Now combine the following information into a single table sorted alphabetically by the name of the ski hill.
```{r, echo = F}
ski_acres
df_cost <- df_cost %>% arrange(ski.resort)
df_comb <- ski_acres %>% bind_cols(df_cost) %>% select(-ski.resort)
str(df_comb)
```

## Solution part 2

```{r}
df_comb <- df_comb %>% mutate(ski.hill = as.character(ski.hill),
                              skiable.acres = as.character(skiable.acres),
                              ticket.cost = as.character(ticket.cost))
str(df_comb)
```

## Solution part 3

```{r}
df_comb <- df_comb %>% rbind(disco) %>% arrange(ski.hill)
df_comb
```

## join()
We could have also used some of the more advanced merge (join) features from dplyr.

```{r}
new.df <- full_join(df1,df2, by='school')
new.df
```

## more about join

The dplyr join command includes the following options:

- left join
- right join
- inner join
- full join
- anti join

## Exercise

Combine the two data sets
```{r,echo = F}
ski_acres
df_cost
disco_df <- tibble(ski.hill = 'Discovery', skiable.acres = '2200', ticket.cost = '20'); disco_df
```

## Solution

```{r}
full_join(ski_acres, df_cost, by = c("ski.hill"= "ski.resort")) %>%
  mutate(ski.hill = as.character(ski.hill),
  skiable.acres = as.character(skiable.acres),
  ticket.cost = as.character(ticket.cost)) %>% 
  full_join(disco_df) %>%
  arrange(ski.hill)
```


# wide / long data

## Hoops Data

This is a _wide_ dataset

```{r, message = F}
hoops <- read_csv('http://www.math.montana.edu/ahoegh/teaching/stat408/datasets/TourneyDetailedResults.csv')
head(hoops)
```

## Plot Creation
Consider a plot where the number of points is colored or faceted by winning or losing. 

This requires a dataset that  is _thin_ and looks something like

|      Game Id     | Points |win/lose | 
|:----------------:|-------|:---------:|
| 1 | 92   | win       |
| 1   | 84   | lose       |
| 2  | 80   | win       |
| 2 | 51    | lose      |

## Data Wrangling

```{r}
lose_score <- hoops %>% select(score = Lscore) %>% 
  mutate(outcome = 'lose', game_id = row_number()) 
win_score <- hoops %>% select(score = Wscore) %>% 
  mutate(outcome = 'win', game_id = row_number())
thin_scores <- bind_rows(win_score, lose_score) %>% 
  arrange(game_id)
thin_scores
```


## Plot Creation

```{r, echo = F, message = F}
thin_scores %>% ggplot(aes(x = score)) + geom_histogram() + facet_wrap(.~outcome) +
  ggtitle('Points scored in NCAA games for Winning and Losing Teams')
```

## tidyr package

The `tidyr` package contains additional tools for converting files from long to wide formats.

Consider the `billboard` dataset which contains the rank of the song (in 2000) for each week after it first entered the list.

```{r}
billboard
```

## billboard data

If we want to identify songs that reach number 1 quickly, the data needs to manipulated.
```{r}
billboard %>% 
  select(artist, track, date.entered, wk1, wk2) %>%
  pivot_longer(
  cols= c('wk1', 'wk2'),
  names_to = 'week',
  values_to = 'rank',
  values_drop_na = T)
```

## Billboard Data Analysis

```{r}
billboard %>% 
  pivot_longer(
  cols= starts_with('wk'),
  names_to = 'week',
  values_to = 'rank',
  values_drop_na = T) %>%
  mutate(week_numb= as.numeric(str_replace(week, 'wk',''))) %>%
  filter(rank == 1) %>%
  arrange(week_numb) %>% 
  slice(1) %>%
  kable()
```

## Exercise

Determine which song in this dataset spent the most time at #1

## Solution

```{r}
billboard %>% 
  pivot_longer(
  cols= starts_with('wk'),
  names_to = 'week',
  values_to = 'rank',
  values_drop_na = T) %>%
  mutate(week_numb= as.numeric(str_replace(week, 'wk',''))) %>%
  filter(rank == 1) %>%
  group_by(track) %>% 
  tally() %>%
  arrange(desc(n)) %>%
  kable()
```


## pivot_wider()

The `pivot_wider()` function allows transformation in the other direction from thin to wide.

# Maps in R

## ggplot functionality
```{r}
mt_counties <- map_data("county", "montana") %>% 
  select(lon = long, lat, group, id = subregion)
class(mt_counties)
mt_counties
```

## Montana Maps
```{r}
mt <- ggplot(mt_counties, aes(lon, lat, group = group)) +
  geom_polygon(fill = "white", colour = "grey50") + 
  coord_quickmap() 
mt
```


## Montana Maps
```{r}
mt + annotate('text',label = 'good guys', x = -111, y = 45.67, col = "darkblue") + 
  annotate('point', x = -111, y = 46, col = "darkblue") +
  annotate('text',label = 'bad guys', x = -114, y = 47, col = "darkred") + 
  annotate('point', x = -114, y = 46.8, col = "darkred")

```

## ggmap

The ggmap package contains nice tools for creating maps; however, it requries creating a google api to fetch the maps.  
[https://github.com/dkahle/ggmap](https://github.com/dkahle/ggmap)

## leaflet
Leaflet is another good option [https://rstudio.github.io/leaflet/](https://rstudio.github.io/leaflet/), however, it does not work well with this html slide conversion.

```{r, echo = F, message = F}
library(leaflet)
seattle_in <- read_csv('http://math.montana.edu/ahoegh/teaching/stat408/datasets/SeattleHousing.csv')
seattle_sort <- seattle_in %>% arrange(price)
m <- leaflet() %>% setView(lng = -122.3, lat = 47.6, zoom = 9)
m  %>% addTiles() %>% addMarkers(lng=seattle_sort$long[c(1, nrow(seattle_sort))], lat=seattle_sort$lat[c(1,nrow(seattle_sort))], popup=c("Cheapest House: 80K ", "Most Expensive House: 7.7 million"))
```

## tmap

tmap also uses the grammar of graphics (gg) principles. [https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)
```{r}
library(tmap)
library(tmap)
data("World")
World

```

## tmap
```{r}
tm_shape(World) +
    tm_polygons("life_exp") 
```

# gg add on packages

## more gg packages

Another of my favorite packages is the ggforce package: [https://github.com/thomasp85/ggforce](https://github.com/thomasp85/ggforce) [https://www.r-bloggers.com/the-ggforce-awakens-again/](https://www.r-bloggers.com/the-ggforce-awakens-again/)

Here is a link to other gg add ons [http://www.ggplot2-exts.org/gallery/](http://www.ggplot2-exts.org/gallery/)

## Exercise

Using either the Seattle Housing dataset or the billboard data, explore the packages above.

## Solution
```{r, echo = F}
library(ggforce)
billboard_thin <- billboard %>% 
  pivot_longer(
  cols= starts_with('wk'),
  names_to = 'week',
  values_to = 'rank',
  values_drop_na = T) %>%
  mutate(week_numb= as.numeric(str_replace(week, 'wk',''))) %>% mutate(date.of.rank = date.entered + (7 * (week_numb - 1)))

nsync <- billboard_thin %>% filter(artist == "N'Sync")

library(ggthemes)
billboard_thin %>% 
ggplot( aes(y = rank, x = date.of.rank, group = track)) + geom_line(alpha = .1) + theme(axis.text.x = element_blank()) + ggtitle('Billboard Song Rankings \n by Week (2000)') + theme_bw() + geom_line(aes(y = rank, x = date.of.rank, group = track, color = track), inherit.aes = F, data = nsync) + theme_wsj()
```

